---
title: "Practical 5: Stationarity Analysis of Commercial Bank Real Estate Loans"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tseries)
library(forecast)
```

**Eshaan Gupta**  
**Roll:** 57  
**Reg. No.:** 230957166


## 1. Question

Suppose we have a data set containing the monthly volume of commercial bank real estate loans (in billions of dollars). Write the R program for the following:

(a) Import data into the R environment.  

(b) Convert the data into a time series object.  

(c) Plot the data to identify the dominant component.  

(d) Check stationarity or non-stationarity using ACF/PACF plot.  

(e) Check stationarity or non-stationarity using the Augmented Dickey-Fuller (ADF) test.  


## 2. Objective

To generate monthly commercial bank real estate loan data, convert it into a time series object, analyze its dominant components visually, and test its stationarity using graphical methods (ACF/PACF) and the Augmented Dickey-Fuller test.


## 3. R Code

```{r code, results='hide', fig.show='hide'}

# Clear workspace
rm(list = ls(all = TRUE))

# Load required libraries
library(tseries)
library(forecast)

# (a) Generate monthly loan data (in billions)
# Simulating 10 years of monthly data (120 observations)

set.seed(123)

trend <- seq(100, 200, length.out = 120)
seasonal <- 10 * sin(2 * pi * (1:120) / 12)
random_noise <- rnorm(120, mean = 0, sd = 5)

loan_data <- trend + seasonal + random_noise

# (b) Convert to time series object
loan_ts <- ts(
  loan_data,
  start = c(2010, 1),
  frequency = 12
)

# (c) Plot the time series
plot(
  loan_ts,
  main = "Monthly Commercial Bank Real Estate Loans",
  ylab = "Loans (in Billions)",
  xlab = "Year",
  col = "blue",
  lwd = 2
)

# (d) ACF and PACF plots
acf(loan_ts, main = "ACF Plot of Loan Series")
pacf(loan_ts, main = "PACF Plot of Loan Series")

# (e) Augmented Dickey-Fuller Test
adf_test <- adf.test(loan_ts)
adf_test
```


## 4. Output

### Time Series Plot

```{r output-plot, echo=FALSE}
plot(
  loan_ts,
  main = "Monthly Commercial Bank Real Estate Loans",
  ylab = "Loans (in Billions)",
  xlab = "Year",
  col = "blue",
  lwd = 2
)
```

### ACF Plot

```{r output-acf, echo=FALSE}
acf(loan_ts, main = "ACF Plot of Loan Series")
```

### PACF Plot

```{r output-pacf, echo=FALSE}
pacf(loan_ts, main = "PACF Plot of Loan Series")
```

### ADF Test Result

```{r output-adf, echo=FALSE}
adf.test(loan_ts)
```


## 5. Conclusion

In this practical, we generated monthly commercial bank real estate loan data and converted it into a time series object with monthly frequency (frequency = 12).

From the time series plot, both trend and seasonal components are visible. The upward movement indicates a dominant trend component, while repeating fluctuations every 12 months indicate seasonality.

The ACF plot shows a slow decay pattern, suggesting non-stationarity. The PACF plot also supports this observation.

The Augmented Dickey-Fuller (ADF) test confirms the stationarity status. If the p-value is greater than 0.05, the series is non-stationary due to trend and seasonality.

Thus, before applying forecasting models, differencing or transformation would be required to make the series stationary.
