---
title: "Practical 8: ARIMA Modeling of AirPassengers Dataset"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tseries)
library(forecast)
library(datasets)
```

**Eshaan Gupta**  
**Roll:** 57  
**Reg. No.:** 230957166


---

## 1. Question

Consider the “AirPassengers” data from R library and write the R program for the following:

(f) Convert the data into a time series object.  
(g) Plot the data to identify the dominant component.  
(h) Decompose the data to observe the dominating components more clearly.  
(i) Check stationarity or non-stationarity using ACF/PACF plot.  
(j) Check stationarity or non-stationarity using the KPSS test.  
(k) If data is non-stationary, make it stationary using an appropriate operator.  
(l) Based on the dominating component, select the suitable technique to fit the data.  
(m) Fit the data using the selected model and estimate the parameters of the model.  
(n) Check the goodness of fit of the model.  

---

## 2. R Code (Question-wise Solution)

```{r}

# To delete previous history or stored variables
rm(list = ls(all = TRUE))

library(tseries)
library(forecast)
library(datasets)


# -------------------------------------------------
# (f) Convert the data into a time series object
# -------------------------------------------------

data(AirPassengers)

ap_data <- as.ts(AirPassengers)

# Printing the data
ap_data



# -------------------------------------------------
# (g) Plot the data to identify dominant component
# -------------------------------------------------

plot(ap_data,
     main = "AirPassengers Data",
     xlab = "Time",
     ylab = "Number of Passengers",
     lwd = 2)



# -------------------------------------------------
# (h) Decompose the data
# -------------------------------------------------

ap_decomp <- decompose(ap_data, type = "multiplicative")
plot(ap_decomp)



# -------------------------------------------------
# (i) Check stationarity using ACF/PACF
# -------------------------------------------------

acf(ap_data,
    main = "ACF of AirPassengers Data",
    lag.max = 20,
    ylim = c(-1,1))

pacf(ap_data,
     main = "PACF of AirPassengers Data",
     lag.max = 20,
     ylim = c(-1,1))



# -------------------------------------------------
# (j) Check stationarity using KPSS Test
# -------------------------------------------------

kpss.test(ap_data)



# -------------------------------------------------
# (k) If non-stationary, make it stationary
# Use log transformation (variance stabilization)
# followed by differencing
# -------------------------------------------------

ap_log <- log(ap_data)

ap_d1 <- diff(ap_log)

kpss.test(ap_d1)

plot(ap_d1,
     main = "First Differenced Log Series",
     xlab = "Time",
     lwd = 2)

acf(ap_d1,
    main = "ACF of Differenced Log Series",
    lag.max = 20,
    ylim = c(-1,1))

pacf(ap_d1,
     main = "PACF of Differenced Log Series",
     lag.max = 20,
     ylim = c(-1,1))



# -------------------------------------------------
# (l) Model Identification using ACF/PACF and AIC
# Since trend and seasonality are dominant,
# we use ARIMA model
# -------------------------------------------------

ap_fit <- auto.arima(ap_log)

ap_fit



# -------------------------------------------------
# (m) Fit the selected model and estimate parameters
# (Model fitted above using auto.arima)
# -------------------------------------------------



# -------------------------------------------------
# (n) Check Goodness of Fit (Diagnostic Checking)
# -------------------------------------------------

res <- ap_fit$residuals

# Plot residuals
plot(res,
     main = "Residuals of AirPassengers Model")

# Check Normality
hist(res,
     main = "Histogram of Residuals")

shapiro.test(res)

# Check Serial Correlation
Box.test(res,
         lag = 20,
         type = "Box-Pierce")

# Check Heteroscedasticity
acf(res^2,
    main = "ACF of Squared Residuals")

pacf(res^2,
     main = "PACF of Squared Residuals")

```

---

## Conclusion

The AirPassengers data shows a strong upward trend and multiplicative seasonality.

ACF and KPSS test confirmed that the original series is non-stationary.

After applying log transformation and first differencing, the series became stationary.

Based on trend and seasonal dominance, an ARIMA model was selected using AIC.

Diagnostic checking confirmed:

- Residuals are approximately normal  
- No serial correlation  
- Constant variance  

Hence, the selected ARIMA model is a good fit for the AirPassengers data.

