---
title: "Practical 6: Time Series Analysis of AirPassengers Dataset"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(datasets)
library(forecast)
library(tseries)
```

**Eshaan Gupta**  
**Roll:** 57  
**Reg. No.:** 230957166


## 1. Question

Consider the “AirPassengers” data from R library and write the R program for the following:

(a) Convert the data into a time series object.  

(b) Plot the data to identify the dominant component.  

(c) Decompose the data to observe the dominating components more clearly.  

(d) Check stationarity or non-stationarity using ACF/PACF plot.  

(e) Check stationarity or non-stationarity using the Augmented Dickey-Fuller (ADF) test.  


## 2. Objective

To analyze the AirPassengers dataset by converting it into a time series object, visualizing its dominant components, decomposing it into trend and seasonal components, and checking its stationarity using graphical and statistical methods.


## 3. R Code

```{r code, results='hide', fig.show='hide'}

# Clear workspace
rm(list = ls(all = TRUE))

# Load required libraries
library(datasets)
library(forecast)
library(tseries)

# (a) Load and convert AirPassengers dataset
data(AirPassengers)

ap_ts <- ts(
  AirPassengers,
  start = c(1949, 1),
  frequency = 12
)

# (b) Plot the time series
plot(
  ap_ts,
  main = "Monthly AirPassengers Data",
  ylab = "Number of Passengers (in thousands)",
  xlab = "Year",
  col = "blue",
  lwd = 2
)

# (c) Decompose the data (Multiplicative Model)
ap_decomp <- decompose(ap_ts, type = "multiplicative")
plot(ap_decomp)

# (d) ACF and PACF plots
acf(ap_ts, main = "ACF Plot of AirPassengers")
pacf(ap_ts, main = "PACF Plot of AirPassengers")

# (e) Augmented Dickey-Fuller Test
adf_test <- adf.test(ap_ts)
adf_test
```


## 4. Output

### Time Series Plot

```{r output-plot, echo=FALSE}
plot(
  ap_ts,
  main = "Monthly AirPassengers Data",
  ylab = "Number of Passengers (in thousands)",
  xlab = "Year",
  col = "blue",
  lwd = 2
)
```

### Decomposition Plot

```{r output-decompose, echo=FALSE}
ap_decomp <- decompose(ap_ts, type = "multiplicative")
plot(ap_decomp)
```

### ACF Plot

```{r output-acf, echo=FALSE}
acf(ap_ts, main = "ACF Plot of AirPassengers")
```

### PACF Plot

```{r output-pacf, echo=FALSE}
pacf(ap_ts, main = "PACF Plot of AirPassengers")
```

### ADF Test Result

```{r output-adf, echo=FALSE}
adf.test(ap_ts)
```


## 5. Conclusion

In this practical, we analyzed the AirPassengers dataset, which contains monthly airline passenger numbers from 1949 to 1960.

The time series plot shows a strong upward trend along with increasing seasonal fluctuations, indicating that both trend and seasonality are dominant components. Since the seasonal variation increases with time, a multiplicative decomposition model was used.

The decomposition clearly separates the trend, seasonal, and irregular components, confirming the presence of strong seasonality and long-term growth.

The ACF plot shows a slow decay pattern, indicating non-stationarity. The PACF plot further supports this observation.

The Augmented Dickey-Fuller (ADF) test typically yields a p-value greater than 0.05 for the original series, confirming that the data is non-stationary.

Therefore, differencing and/or transformation (such as logarithmic transformation) would be required before fitting forecasting models like ARIMA.
