---
title: "Practical 9: Forecast of commercial loans"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(forecast)
```

**Eshaan Gupta**  
**Roll:** 57  
**Reg. No.:** 230957166


## 1. Question

Suppose we have a data set containing the monthly volume of commercial bank real estate loans(in billions of dollars) stored in a text file. Write the R code to forecast the volume of commercial loans for next 20 months.


## 2. Objective

The primary goal of this analysis is to quantify and project the future trajectory of commercial bank real estate loans based on historical monthly volume data.


## 3. Code

```{r code, results='hide', fig.show='hide'}
# Set working directory (edit path as needed)
setwd("D:\\Eshaan\\TSA Lab\\TSA-LAB-main")


# Import data from a text file
raw_data <- read.table("bank_case.txt", header = FALSE)

# 3. Convert to a Time Series (ts) object
# Change 'start' and 'frequency' based on your data's actual start date
# frequency = 12 indicates monthly data
loan_ts <- ts(raw_data$V1, start = c(2020, 1), frequency = 12)

# 4. Fit an automated forecasting model (Auto-ARIMA)
# auto.arima automatically selects the best parameters for trend and seasonality
fit <- auto.arima(loan_ts)

# 5. Forecast the next 20 months
loan_forecast <- forecast(fit, h = 20)

# 6. View and Plot the results
print(loan_forecast)

```

## 4. Plot

```{r output-plot, echo=FALSE}
loan_forecast
plot(loan_forecast, 
     main = "20-Month Forecast: Commercial Bank Real Estate Loans",
     ylab = "Billions of Dollars", 
     xlab = "Year",
     col = "blue")
```

## 5. Conclusion

Based on the time-series analysis and the application of the Auto-Regressive Integrated Moving Average (ARIMA) model, the following conclusions can be drawn regarding the commercial bank real estate loan volume:

Forecasted Trend: The model identifies a consistent increasing trend in loan volumes. Over the next 20 months, the total volume is expected to reach approximately $120.92885 billion by the end of the period.

Statistical Confidence: The forecast provides a high degree of certainty in the short term; however, the widening of the 80% and 95% confidence intervals over the 20-month horizon indicates that external economic volatility remains a significant factor for long-term accuracy.

Market Implication: If the current trajectory continues, the commercial real estate sector will likely see a expansion of credit, suggesting that banking institutions are currently operating under a growth-oriented lending posture relative to historical data.
