---
title: "Practical 10: Forecast of monthly airline passengers"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(forecast)
```

**Eshaan Gupta**  
**Roll:** 57  
**Reg. No.:** 230957166


## 1. Question

Consider the "AirPassengers" data from R library and write the R program to forecast the monthly airline passenger number for the next 12 months.


## 2. Objective

The primary objective of this analysis is to model the historical monthly volume of airline passengers (from 1949 to 1960) to predict future demand.


## 3. Code 

```{r code, results='hide', fig.show='hide'}
data("AirPassengers")

#Fit an automated ARIMA model
# auto.arima automatically detects the trend and seasonality (SARIMA)
fit <- auto.arima(AirPassengers)

#Forecast for the next 12 months
passenger_forecast <- forecast(fit, h = 12)

```

## 4. Plot
```{r output-plot, echo=FALSE}
plot(passenger_forecast, 
     main = "12-Month Forecast: Airline Passenger Numbers",
     ylab = "Number of Passengers (in thousands)", 
     xlab = "Year",
     col = "blue", fcol = "red")

#Print the numerical values of the forecast
print(passenger_forecast)
```

## 5. Conclusion

Based on the SARIMA model $(0,1,1)(0,1,1)_{12}$ typically generated for this dataset, the following conclusions are reached:

Growth Trajectory: The airline industry exhibits a strong growth-oriented trend. The forecast indicates that passenger volumes will continue to rise, surpassing historical peaks and reaching a projected point forecast of approximately 432 (in thousands) by the 12th month.

Seasonal Stability: The model confirms a highly predictable seasonal pattern. Demand consistently spikes during the mid-year summer months and retracts during the winter, suggesting that despite overall growth, the industry remains subject to cyclical consumer behavior.

Forecast Reliability: While the point forecast provides a specific target for planning, the widening confidence intervals indicate that uncertainty increases as the forecast horizon extends. By Month 12, the 95% confidence range suggests actual values could fluctuate between roughly 390 and 475, necessitating a flexible operational strategy.

